Long term cloud enhancement events of global solar irradiance over Thessaloniki

### 2024-06-03 article/article.Rmd
Commit 2024-06-03 13:10

author:
  - name:        Athanasios N. Natsis
    email:       natsisphysicist@gmail.com
    affiliation: LAP
    correspondingauthor: true
    footnote: 1
  - name:        Alkiviadis Bais
    email:       abais@auth.gr
    affiliation: LAP
  - name:        Charikleia Meleti
    email:       meleti@auth.gr
    affiliation: LAP
    footnote: 2
address:
  - code:         LAP
    organization: Laboratory of Atmospheric Physics, Physics Department, Aristotle University of Thessaloniki
    # addressline:  1 main street
    city:         Thessaloniki
    postcode:     54124
    country:      Greece
  - code:         Another University
    organization: Department
    addressline:  A street 29
    postcode:     2054 NX
    city:         Manchester,
    state:        State
    country:      The Netherlands
footnote:
  - code: 1
    text: "This is the first author footnote."
  - code: 2
    text: "Another author footnote."
abstract: |
  Here wii be the abstract.

  It consists of two paragraphs.

@@

keywords: 
  - cloud enhancement
  - total solar radiation
  - global horizonta irradiance
  - over irradiance

journal:        "An awesome journal"
date:           "`r Sys.Date()`"
linenumbers:    false
numbersections: true
bibliography:   bibliography.bib

biblio-style:   elsarticle-harv # author year style for natbib - use 'elsarticle-num' or 'elsarticle-num-names' for numbered scheme

classoption:    preprint, 3p, authoryear # remove authoryear is not using `elsarticle-harv`
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/elsevier-harvard

@@

```{r setup, echo=F, include=FALSE}
rmarkdown::find_pandoc(dir = '/usr/lib/rstudio/resources/app/bin/quarto/bin/tools')
# renv::load("~/MANUSCRIPTS/02_enhancement/", quiet = TRUE)
## document configurations
knitr::opts_chunk$set(comment    = ""      )
# knitr::opts_chunk$set(dev        = "pdf"   )
knitr::opts_chunk$set(out.width  = "50%"   )
knitr::opts_chunk$set(fig.align  = "center")
# knitr::opts_chunk$set(fig.pos    = 'h!'    )
options(knitr.graphics.auto_pdf = FALSE)
library(pander)
library(data.table)
library(kableExtra)
library(dplyr)
## load data from trends analysis
source("~/MANUSCRIPTS/02_enhancement/GHI_enh_00_variables.R")
load("~/MANUSCRIPTS/02_enhancement/data/GHI_enh_03_process.Rda")
TSIinfo <- data.table(read.csv("~/MANUSCRIPTS/02_enhancement/figures/tbl_tsi_info.dat",
                               strip.white = TRUE,
                               sep = ";",
                               header = TRUE))
```

@@

\  
\  
CE: cloud enhancement cases one minute

@@

Our dataset, after the data selection processing, consists of
`r ST_total$wattGLB.N` records of GHI, of which
$`r round(100*ST_total$GLB_diff.N_pos/ST_total$wattGLB.N, 3)`\,\%$ are CE events and
$`r round(100*ST_extreme_total$GLB_diff.N_pos/ST_total$wattGLB.N, 3)`\,\%$ are ECE events.
The highest GHI recorder was 
$`r round(ST_daily[which.max(wattGLB.max), wattGLB.max],1)`\,W/m^2$
on `r strftime(ST_daily[which.max(wattGLB.max), Date], "%d\\ %B %Y")`.
The absolute stronger CE event had an OIR of 
$`r round(ST_daily[which.max(GLB_diff.max), GLB_diff.max], 2)`\,W/m^2$ on
`r strftime(ST_daily[which.max(GLB_diff.max), Date], "%d\\ %B %Y")`.
The relative stronger CE event was 
$`r round(100*ST_daily[which.max(GLB_ench.max), GLB_ench.max],1)`\,\%$ above the 
clear sky threshold, on 
`r strftime(ST_daily[which.max(GLB_ench.max), Date], "%d\\ %B %Y")`.

## Trends

@@

```{r P-energy-N, echo=FALSE, fig.cap="Trend of yearly CE number of occurancies."}
knitr::include_graphics("../images/P_energy-6.png")
```

@@

## CE Climatology

Another interesting aspect of the CE cases occurrences, is their seasonal cycle. In
Figure\nobreakspace{}\ref{fig:relative-month-occurancies}, we have a monthly box plot
(whisker plot), where the values have been normalized by the highest median value,
that occurs on June.  Although the number of occurrences has a wide spread throughout
the study period, the most active period of CE occurrences is during May and June.
During the Winter (December -- February) the CE cases are about $25\,\%$ of the maximum.
For the rest months the occurrences gradually ramp between the maximum and minimum.
These seasonality is a combinations of the clouds occurrences, and the relevant
position of the sun during the year. Unfortunately, the lack of detailed data of cloud
formation, type and location is not allowing further analysis.

```{r relative-month-occurancies, echo=FALSE, fig.cap="Seasonal statistics of the number of CE occurancies for each month, normalized to the maximum occurancies on June. The box represents the values of the low $25\\,\\%$ percentile to $75\\,\\%$ percentile, where the thick horizontal line inside is the mean, the verical lines extend to the macimum and minimum vales, the dots are outlier values, and the rhombus is the mean."}
knitr::include_graphics("../images/clim_CE_month_norm_MAX_median_N-2.png")
```
<!--
```{r echo=FALSE, fig.cap="Distribution of the number of CE cases"}
knitr::include_graphics("../images/climCEmonth-1.png")
```
--->

@@

The relation between the duration and the mean OIR intensity of the groups have been
studied (Figure\nobreakspace{}\ref{fig:group-2d}). We observe that the GCE events
tend to have either long duration, or large intensity, events with strong enhancement
and long duration are very rare. A similar result on this relation, have been also
reported by @Zhang2018, on a study using a far higher sampling rate than ours.

```{r group-2d, echo=FALSE, fig.cap="Relation of mean over irradiance and CE group duration"}
knitr::include_graphics("../images/P-groups-bin2d-1.png")
```

@@

## Extreme CE above TSI

A characteristic of the CE events that is commonly reported and has some significance
on the solar energy production infrastructure are the extreme CE events (ECE). Where
solar irradiance exceeds the expected irradiance on top of the atmosphere, we have
defined this as ECE (Equation\nobreakspace{}\ref{eq:ECE}).

<!-- -  max value ECE -->
<!--   relative to other works max values -->

Analogous to Figure\nobreakspace{}\ref{fig:relative-month-occurancies} we have
computed the distribution of the number of occurrences of ECE events by month in
Figure\nobreakspace{}\ref{fig:relative-month-occurancies-ECE}. The most active period
for ECE events is in the spring and the start of the summer (March -- June), followed
by a period in the late fall (September and October).  This is probably an effect of
the weather characteristics of those periods where, there are continuous alternations
between clear sky periods and clouds.

```{r relative-month-occurancies-ECE, echo=FALSE, fig.cap="Seasonal statistics of the number of ECE occurancies for each month, normalized to the maximum occurancies on March. The box represents the values of the low $25\\,\\%$ percentile to $75\\,\\%$ percentile, where the thick horizontal line inside is the mean, the verical lines extend to the macimum and minimum vales, the dots are outlier values, and the rhombus is the mean."}
knitr::include_graphics("../images/clim_ECE_month_norm_MAX_median_N-2.png")
```

@@

The distribution of the ECE events (Figure\nobreakspace{}\ref{fig:P-extreme-distribution}), shows that there are rare cases where the OIR can exceed the 
TSI even more than $400\,\text{W}/\text{m}^2$, with the $75\,\%$ of the cases to be below $200\,\text{W}/\text{m}^2$. Similar to CE cases, the  distribution of ECE have a similar pattern
with the results of @Vamvakas2020, with the difference that the OIR values are lower fro Thessaloniki.

@@

By creating a clear sky approximation of the GHI, which can represent the long- and
short-term variation of the expected clear sky GHI, we were able to identify cases of 
CE events.
By analyzing the CE cases we found an increase of 
$`r round(yearlyCE[var == "GLB_diff.N", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$,
with the
mean annual total energy of the CE events increasing with a rate of
$`r round(yearlyCE[var == "GLB_diff.sum", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kj}/\text{year}$.
The most active season of CE events over Thessaloniki is concentrated on early
summer, on May and June.

The magnitude of the ECE events doesn't exceeds the values reported from sites with
more favourable conditions for the phenomenon. Although, the climatological
characteristic of the ECE events, showed that the most active months are spread on
half of the year (March -- June and September and October).
We found that CE conditions, can have a duration of more than an hour in rare
cases, with the bulk of the cases having a duration under 5 minutes.
Some of the characteristics of CE and ECE events we analysed,
have strong similarities with results
by @Vamvakas2020, for a city southern of Thessaloniki, with analogues differences, due to the 
 higher intensity of the solar radiation.

An interpretation of CE trends, shows that the interaction of GHI with the clouds
through this 30 year period, is a dynamic phenomenon that's needs further investigation.
Although, to approach it, we need more long term observations of AOD and clouds
characteristics.

@@

```{r example-day, echo=FALSE, fig.cap="Example of cloud indentification data for 2019-07-11. Where are denoted with red circles all the detected CE events, the cloud intearaction with the GHI (blue crosses), and the thesholds of CE and the threshold of ECE events with red and black line respectively."}
knitr::include_graphics("../images/example_days-18.png")
```

```{r example-year, echo=FALSE, fig.cap="Example of the relation of the measured GHI versus the reference clear sky irradiance for the year 2005, where the intesity of the detected over irradianse (OIR) is color coded."}
knitr::include_graphics("../images/example_years-13.png")
```

@@

<!--
Elsevier template

Please make sure that your manuscript follows the guidelines in the 
Guide for Authors of the relevant journal. It is not necessary to 
typeset your manuscript in exactly the same way as an article, 
unless you are submitting to a camera-ready copy (CRC) journal.

For detailed instructions regarding the elsevier article class, see   <https://www.elsevier.com/authors/policies-and-guidelines/latex-instructions>

# Bibliography styles

Here are two sample references: @Feynman1963118 [@Dirac1953888].

By default, natbib will be used with the `authoryear` style, set in `classoption` variable in YAML and with `elsearticle-harv.bst` which is among provided style by `elsarticle` documentclass. Other available style are `elsarticle-num.bst` and `elsarticle-num-names.bst` — the first one can be used for the numbered scheme, second one for numbered with new options of natbib.sty.

You can sets extra options with `natbiboptions` variable in YAML header. Example 
```yaml
natbiboptions: longnamesfirst,angle,semicolon
```

There are various more specific bibliography styles available at
<https://support.stmdocs.in/wiki/index.php?title=Model-wise_bibliographic_style_files>. 
To use one of these, add it in the header using, for example, `biblio-style: model1-num-names`.

## Using CSL

If `citation_package` is set to `default` in `elsevier_article()`, then pandoc is used for citations instead of `natbib`. In this case, the `csl` option is used to format the references. Alternative `csl` files are available from <https://www.zotero.org/styles?q=elsevier>. These can be downloaded
and stored locally, or the url can be used as in the example header.

# Equations

Here is an equation:
$$ 
  f_{X}(x) = \left(\frac{\alpha}{\beta}\right)
  \left(\frac{x}{\beta}\right)^{\alpha-1}
  e^{-\left(\frac{x}{\beta}\right)^{\alpha}}; 
  \alpha,\beta,x > 0 .
$$

Here is another:
\begin{align}
  a^2+b^2=c^2.
\end{align}

Inline equations: $\sum_{i = 2}^\infty\{\alpha_i^\beta\}$

# Figures and tables

Figure \ref{fig2} is generated using an R chunk.

```{r fig2, fig.width = 5, fig.height = 5, fig.align='center', out.width="50%", fig.cap = "\\label{fig2}A meaningless scatterplot.", echo = FALSE}
plot(runif(25), runif(25))
```

# Tables coming from R

Tables can also be generated using R chunks, as shown in Table \ref{tab1} for example.

```{r tab1, echo = TRUE}
knitr::kable(head(mtcars)[,1:4], 
    caption = "\\label{tab1}Caption centered above table"
)
```

-->

>>>>

author:
  - name:        Athanasios N. Natsis
    email:       natsisphysicist@gmail.com
    affiliation: LAP
    correspondingauthor: true
    # footnote: 1
  - name:        Alkiviadis Bais
    email:       abais@auth.gr
    affiliation: LAP
  - name:        Charikleia Meleti
    email:       meleti@auth.gr
    affiliation: LAP
    # footnote: 1
address:
  - code:         LAP
    organization: Laboratory of Atmospheric Physics, Physics Department, Aristotle University of Thessaloniki
    # addressline:  1 main street
    city:         Thessaloniki
    postcode:     54124
    country:      Greece
  # - code:         Another University
  #   organization: Department
  #   addressline:  A street 29
  #   postcode:     2054 NX
  #   city:         Manchester,
  #   state:        State
  #   country:      The Netherlands
footnote:
  - code: 1
    text: "This is the first author footnote."
  - code: 2
    text: "Another author footnote."
abstract: |
  Here will be the abstract.

@@

keywords: 
  - cloud enhancement
  - total solar radiation
  - global horizontal irradiance
  - over irradiance

journal:        "Atmospheric Research"
date:           "`r Sys.Date()`"
linenumbers:    false
numbersections: true
bibliography:   bibliography.bib

biblio-style:   elsarticle-num-names # elsarticle-harv author year style for natbib - use 'elsarticle-num' or 'elsarticle-num-names' for numbered scheme

classoption:    preprint, 5p, authoryear # remove authoryear is not using `elsarticle-harv`
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/elsevier-harvard

@@

```{r setup, echo=F, include=FALSE}
rmarkdown::find_pandoc(dir = '/usr/lib/rstudio/resources/app/bin/quarto/bin/tools')
# renv::load("~/MANUSCRIPTS/02_enhancement/", quiet = TRUE)
## document configurations
knitr::opts_chunk$set(comment    = ""      )
# knitr::opts_chunk$set(dev        = "pdf"   )
knitr::opts_chunk$set(out.width  = "100%"   )
knitr::opts_chunk$set(fig.align  = "center")
# knitr::opts_chunk$set(fig.pos    = 'h!'    )
options(knitr.graphics.auto_pdf = FALSE)
library(pander)
library(data.table)
library(kableExtra)
library(dplyr)
## load data from trends analysis
source("~/MANUSCRIPTS/02_enhancement/GHI_enh_00_variables.R")
load("~/MANUSCRIPTS/02_enhancement/data/GHI_enh_03_process.Rda")
TSIinfo <- data.table(read.csv("~/MANUSCRIPTS/02_enhancement/figures/tbl_tsi_info.dat",
                               strip.white = TRUE,
                               sep = ";",
                               header = TRUE))
```

@@

\  
\

CE: cloud enhancement cases one minute

@@

Our dataset, after the data selection processing, consists of
`r ST_total$wattGLB.N` records of GHI measurements, of which
$`r round(100*ST_total$GLB_diff.N_pos/ST_total$wattGLB.N, 3)`\,\%$ were identified as CE events and
$`r round(100*ST_extreme_total$GLB_diff.N_pos/ST_total$wattGLB.N, 3)`\,\%$ as ECE events.
The highest recorded GHI due to CE was
$`r round(ST_daily[which.max(wattGLB.max), wattGLB.max],1)`\,\text{W}/\text{m}^2$
on `r strftime(ST_daily[which.max(wattGLB.max), Date], "%d\\ %B %Y")`, and had
the maximum value of OIR at $`r round(ST_daily[which.max(GLB_diff.max), GLB_diff.max], 1)`\,\text{W}/\text{m}^2$.
<!-- The absolute strongest CE event had an OIR of 
<!-- $`r round(ST_daily[which.max(GLB_diff.max), GLB_diff.max], 2)`\,\text{W}/\text{m}^2$ on
<!-- `r strftime(ST_daily[which.max(GLB_diff.max), Date], "%d\\ %B %Y")`. -->
The relative stronger CE event was 
$`r round(100*ST_daily[which.max(GLB_ench.max), GLB_ench.max],1)`\,\%$ above the 
clear sky threshold, on 
`r strftime(ST_daily[which.max(GLB_ench.max), Date], "%d\\ %B %Y")`.

## Long term trends

@@

```{r P-energy-N, echo=FALSE, fig.cap="Trend of yearly cloud enhancement occurrences."}
knitr::include_graphics("../images/P_energy-6.png")
```

@@

## Cloud enhancements climatology

Another interesting aspect of the CE cases occurrences, is their seasonal cycle. In
Figure\nobreakspace{}\ref{fig:relative-month-occurrences}, we have the monthly box plot
(whisker plot), where the values have been normalized by the highest median value,
that occurs on June.  Although the number of occurrences has a wide spread throughout
the study period, the most active period of CE occurrences is during May and June.
During the Winter (December -- February) the CE cases are about $25\,\%$ of the maximum.
For the rest months, the occurrences gradually ramp between the maximum and minimum.
This seasonality is a combinations of the clouds occurrences, and the relevant
position of the sun during each season of the year. Unfortunately, the lack of detailed data on cloud
formation, type and location is not allowing further analysis.

```{r relative-month-occurrences, echo=FALSE, fig.cap="Seasonal statistics of the number of CE occurrences for each month, normalized to the maximum occurrences on June. The box represents the values of the low $25\\,\\%$ percentile to $75\\,\\%$ percentile, where the thick horizontal line inside is the mean, the vertical lines extend to the maximum and minimum values, the dots are outlier values, and the rhombus is the mean."}
knitr::include_graphics("../images/clim_CE_month_norm_MAX_median_N-2.png")
```
<!--
```{r echo=FALSE, fig.cap="Distribution of the number of CE cases"}
knitr::include_graphics("../images/climCEmonth-1.png")
```
--->

@@

The relation between the duration and the mean OIR intensity of the groups have also been
studied (Figure\nobreakspace{}\ref{fig:group-2d}). We observed that the GCE events
tend to have either long duration, or large intensity. Events with strong enhancement
and long duration are very rare. Similar results on this relation, have been
reported by @Zhang2018, on a study using a far higher sampling rate than ours.

```{r group-2d, echo=FALSE, fig.cap="Relation of the mean over irradiance and duration of GCE, where the color scale denotes the frequency of the respected events."}
knitr::include_graphics("../images/P-groups-bin2d-1.png")
```

@@

## Extreme cloud enhancement events

An aspect of the CE events that is commonly reported and has some significance
on the solar energy production infrastructure are the extreme CE events (ECE). Where
solar irradiance exceeds the expected irradiance on top of the atmosphere
(Equation\nobreakspace{}\ref{eq:ECE}).
Analogous to Figure\nobreakspace{}\ref{fig:relative-month-occurancies} we have
computed the distribution of the number of occurrences of ECE events by month in
Figure\nobreakspace{}\ref{fig:relative-month-occurancies-ECE}. The most active period
for ECE events is in the spring and the start of the summer (March -- June), followed
by a period in the late fall (September and October).  This is probably related to 
the weather characteristics of these periods, where there are continuous alternations
between clear sky periods and clouds.

```{r relative-month-occurancies-ECE, echo=FALSE, fig.cap="Seasonal statistics of the number of ECE occurancies for each month, normalized to the maximum occurancies on March. The box represents the values of the low $25\\,\\%$ percentile to $75\\,\\%$ percentile, where the thick horizontal line inside is the mean, the vertical lines extend to the maximum and minimum vales, the dots are outlier values, and the rhombus is the mean."}
knitr::include_graphics("../images/clim_ECE_month_norm_MAX_median_N-2.png")
```

@@

The distribution of the ECE events (Figure\nobreakspace{}\ref{fig:P-extreme-distribution}),
shows that there are rare cases where the OIR can exceed the TSI even more than
$400\,\text{W}/\text{m}^2$, with the $75\,\%$ of the cases to be below
$200\,\text{W}/\text{m}^2$. 
Those finds are in accordance with results form @Vamvakas2020, 
with the difference that the OIR values are lower for Thessaloniki.

@@

By creating a clear sky approximation of the GHI, which represents the long- and
short-term variation of the expected clear sky GHI, we were able to identify cases of 
CE events.
After analyzing the CE cases, we found an increase of 
$`r round(yearlyCE[var == "GLB_diff.N", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$,
with the
mean annual total energy of the CE events increasing with a rate of
$`r round(yearlyCE[var == "GLB_diff.sum", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kj}/\text{year}$.
The most active season of CE events over Thessaloniki is concentrated on early
summer, on May and June.

The magnitude of the ECE events doesn't exceeds the values reported from sites with
more favourable conditions for the phenomenon. Although, the climatological
characteristic of the ECE events, showed that the most active months are spread on
half of the year (March -- June and September and October).
We found that CE conditions, can have a duration of more than an hour in rare
cases, with the bulk of the cases having a duration under 5 minutes.
Some of the characteristics of CE and ECE events we analysed,
have strong similarities with results
by @Vamvakas2020, for a city southern of Thessaloniki, with the analogues differences on the
 intensity of the solar radiation.

An interpretation of the CE trends, shows that the interaction of GHI with the
clouds,
through this 30 year period, is a dynamic phenomenon that's needs further investigation.
Although, to approach it, we need more long term observations of AOD and clouds
characteristics.

@@

```{r example-day, echo=FALSE, fig.cap="Example of cloud identification data for 2019-07-11. Where are denoted with red circles all the detected CE events, the cloud interaction with the GHI (blue crosses), and the thresholds of CE and the threshold of ECE events with red and black line respectively.", fig.pos='H'}
knitr::include_graphics("../images/example_days-18.png")
```

```{r example-year, echo=FALSE, fig.cap="Example of the relation of the measured GHI versus the reference clear sky irradiance for the year 2005, where the intensity of the detected over irradiance (OIR) is color coded.", fig.pos='H'}
knitr::include_graphics("../images/example_years-13.png")
```

@@
